---
title: "Impact of Texas Prison Expansion On Teen Pregnancy"
author: "Patrick Chase Anuka Revi"
date: "4/21/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =FALSE,
                      include=FALSE,
                      message=FALSE,
                      warning=FALSE)
```

```{r, load libraries}
library(tidyverse)
library(haven)
library(Synth)
library(devtools)
if(!require(SCtools)) devtools::install_github("bcastanho/SCtools")
library(SCtools)
library(dplyr)
library(kableExtra)
library(skimr)
library(lubridate)
library(ggcorrplot)
```









# **Introduction**




In this project we are looking at the impact of prison expansion on teenage pregnancy and live birth rate. Specifically , we are focusing on Texas state prison expansion during 1993 and early 2000 implemented by governor Ann Richards that tripled the size of the states prison system. Texas sudden prison boom had a major  consequence – state’s prison population more than doubled in five years. It would be reasonable to assume that since many young men, especially young black men  were incarcerated, teen fertility would drop due to imbalance in sex ratio. The goal of this paper is to determine if there is a causal relationship between the spike of incarceration rate among juveniles and teen pregnancy and live birth rate. We would expect this relationship to be negative, in order to evaluate this we are using synthetic controls. The advantage of using SCM is it looks like the treated case in key metrics in terms of the prior covariates and other post outcome predictors, thus it gives us apples to apples comparison. 

 
 


# **Data Exploration**



We used *Texas.dta* data found in Dr. Scott Cunningham's textbook *"Mixtape"* <https://mixtape.scunning.com/synthetic-control.html?panelset=r-code> and merged it with teen pregnancy data from CDC.gov. <https://www.cdc.gov/nchs/data-visualization/teen-births/index.htm>
Data overview is given in the table below. Our variable of interest is teen pregnancy given by the variable name *"state.Births"* (in absolute terms) or *"state rate"*(state birth per 1000). 







```{r Data load and clean, error=FALSE}


# data read in 
teen_preg <- read.csv("https://raw.githubusercontent.com/patrick-chase/Hello-World/main/Causal%20Inference%20Working%20Folder/NCHS_-_U.S._and_State_Trends_on_Teen_Births.csv")

read_data <- function(df)
{
  full_path <- paste("https://raw.github.com/scunning1975/mixtape/master/", 
                     df, sep = "")
  df <- read_dta(full_path)
  return(df)
}

texas <- read_data("texas.dta") %>%
  as.data.frame(.) 

# Filtering data sets for 1990-2000 and focusing on all teenagers (Age.Group..Years == "18-19 years")
teen_preg <- teen_preg %>%
  filter(ï..Year >= 1990,ï..Year <= 2000) %>%
  rename(state = State, year = ï..Year)



teen_preg <- subset(teen_preg, state != "Total U.S." & Age.Group..Years. != "15-17 years" & Age.Group..Years. != "18-19 years")

texas <- texas %>%
  filter(year >= 1990, year <= 2000)%>%
  group_by(year) 

#merging data sets
prep_data_in <- merge(teen_preg, texas, by = c("year", "state"))
prep_data_in$State.Births <-  as.numeric(as.character(prep_data_in$State.Births))

prep_data_in_try<-prep_data_in%>%rename(teen_births=)

```


<bk><bk><bk>


```{r, include=TRUE}
head(prep_data_in)%>%kable()
```


<bk><bk><bk>


It is important to visualize how teen pregnancy & incarceration  rate have changed over years. Figure 1 below shows 3 graphs, first one is an average teen pregnancy rate which has been decreasing from 1991 till 2000. We do notice a decrease at a slower rate around the year of 1997 but the overall trend is obvious. On the same figure we see average number of incarcerated black and white men that has been increasing from 1990. The general trend however looks very different for these different groups, white male incarceration decreased from 1993 till 1995 and from 1997 till 1999, while incarceration rate for black men has been increasing consistently.


Correlation Table below shows that our data suffers from multicollinearity. We have some variables such as income and probation that has strong correlation with our variable of interest. In the next section, we will perform synthetic analysis in an attempt to identify causal relationship between teenage pregnancy and incarceration rate in the US from 1980-2000. 



 



```{r, fig.cap="Figure1", include=TRUE, fig.align="center"}
data<-prep_data_in%>%
 mutate(yr=as.Date(as.character(year), format="%Y"),
        y=year(yr))


g2<-data%>%group_by(y)%>%
  summarize(avg_birth=mean(State.Rate, na.rm=TRUE))%>%
  ggplot(aes(y,avg_birth))+geom_line(color="blue")+theme(axis.text.x=element_text(angle=60, hjust=1))+xlab("Average Number of Teenage Pregnancy")


g3<-data%>%group_by(y)%>%
  summarize(avg_incarbl=mean(bmprison, na.rm=TRUE))%>%
  ggplot(aes(y,avg_incarbl))+geom_line(color="blue")+theme(axis.text.x=element_text(angle=60, hjust=1))+xlab("Incarcerated Black Men")



g4<-data%>%group_by(y)%>%
  summarize(avg_incarwh=mean(wmprison, na.rm=TRUE))%>%
  ggplot(aes(y,avg_incarwh))+geom_line(color="blue")+theme(axis.text.x=element_text(angle=60, hjust=1))+xlab("Incarcerated White Men")

source("http://peterhaschke.com/Code/multiplot.R")
multiplot(g2,g3, g4, cols=3)

```




<bk><bk><bk>



```{r, include=TRUE, out.width="90%", fig.align="center"}
num_cols <- unlist(lapply(prep_data_in, is.numeric)) # Identify numeric columns
data_num <- prep_data_in[, num_cols]   
data_num<-na.omit(data_num)


corr<-cor(data_num,use="pairwise.complete.obs")%>%round(2)


ggcorrplot(corr, hc.order = FALSE,
           lab = TRUE, 
           lab_size =3,
           tl.cex=8,
           method="square",
           type = "lower",
           colors = c("tomato2", "white", "springgreen3"), 
           title="Correlogram of birthweight variables", 
           ggtheme=theme_bw)
```


\newpage



# **Synthetic Control - Literature Review**


Synthetic control methods were originally proposed in Abadie and Gardeazabal (2003) and
Abadie et al. (2010) with the aim to estimate the effects of aggregate interventions, that is,
interventions that are implemented at an aggregate level affecting a small number of large units
(such as a cities, regions, or countries), on some aggregate outcome of interest.
Synthetic control uses data driven procedure and  is based on the idea that, when the units of observation are
a small number of aggregate entities, a combination of unaffected units often provides a more appropriate comparison than any single unaffected unit alone. SCM is transparent as we can tell how well synthetic control's outcome matches with the treated region's outcome before the treatment (some type of policy intervention).




To create a synthetic control, data must be available for several periods before the treatment for
the policy region and the pool of potential donor regions.
Some assumptions in order for SCM to estimate unbiased effects are: i) policy (treatment) in the affected region cannot affect the outcomes in the pool of donor regions; ii)variables used to form the weights must have the donor pool regions that are similar to those for the affected region; iii) Variables and the outcome must have an approximate linear relationship.



The SCM received much attention following Abadie, Diamond, and
Hainmueller (2010)—hereafter, ADH— and the code’s public release. In that article, the
authors examine how California’s tobacco control program under Proposition 99, implemented in 1988,
affected smoking by creating a synthetic control version of California. They estimate that by 2000, per
capita sales of cigarette packs had fallen 26 packs because of the program. ADH assumed that the treatment had no effect on the outcome variable before the 1988 treatment
(e.g., people did not stockpile cigarette packs in anticipation of the law’s passage).
t. ADH also assumed that the outcomes were
independent across states, implying that the treatment in California did not affect the outcome in donor
pool states. There must be states with predictor
values near those of the treated state before treatment. ADH illustrate this with a problematic example:
if one explanatory variable is the share of the population that is white, and California is 80 percent
white, it could be represented by a synthetic state formed from an equal weighting of a 65 percent
white/35 percent nonwhite state with a 95 percent white/5 percent nonwhite state. But if a strong
nonlinear relationship exists between racial composition and smoking, the synthetic state could poorly
approximate the treated state, resulting in interpolation bias. Abadie, Diamond, and Hainmueller (2015)
also declare that the outcomes of the states in the donor pool should be driven by the same process as
that found in the treated state before treatment.



\newpage


## **Synthetic Control Method**


->Step 1: Identify predictors of the outcome variable


->Step 2: : Identify possible donor states to synthesize the control state.


->Step 3: Choose a method for selecting predictor weights.


->Step 4: Assess the pretreatment period goodness of fit of the synthetic control state (generated
using the Synth package).


->Step 5: Conduct placebo test on states in the donor pool to evaluate the significance of the results
for the treated state.


->Step 6: Conduct sensitivity analyses to further test the credibility of the results.






\newpage



# **Synthetic Control for Teen Pregnancy - Analysis**




To construct our synthetic control unit we define a vector of weights **W**. Each **W** then represents one particular weighted average of control units and therefore one potential synthetic control unit.To create the most similar synthetic control unit, the synth() function chooses the vector W^* to minimize a distance, kX_{1} − X_{0}Wk
, between **X_{1}** and **X_{0}W**, subject to the weight constraints. 


The **V** matrix is introduced to allow different weights to the variables in X_{0} and X_{1} depending on their predictive power on the outcome. An optimal choice of **V** assigns weights that minimize the mean square error of the synthetic control estimator, that is the expectation of (Y_{1} − Y_{0}W)'(Y_{1} − Y_{0}W)



In this procedure a V* is chosen among all positive definite and diagonal matrices such that the mean squared prediction error (MSPE) of the outcome variable is minimized over some set of pre-intervention periods.



Abadie et al. (2010) describe how synthetic control methods facilitate inferential techniques akin to permutation tests-so-called placebo studies - iteratively apply the synthetic control method by randomly reassigning the intervention in time (i.e., pre-intervention dates)or across units (i.e., to control units where the intervention did not occur) to produce a set of placebo effects. Subsequently, we can compare the set of placebo effects to the effect that was estimated for the time and unit where the intervention actually occurred. This comparison is informative about the rarity of the magnitude of the treatment effect that was observed for the exposed unit.



The first step is to reorganize the panel data set into an appropriate format that is suitable for the main estimator function synth(). At a minimum, synth() requires as inputs the four data matrices X_{1}: control cases before the treatment (in our case teen pregnancy rate in every other state except Texas before the prison expansion), X_{0} : the control cases after the treatment(teen pregnancy rate in every other state after the expansion except Texas), Z_{1}: the treatment case before the treatment (teen pregnancy rate in Texas state before the prison expansion), and Z_{0}: treatment case after the treatment (teen pregnancy rate in Texas state after the prison expansion ) that are needed to construct a synthetic control unit. The corresponding R code is given below:






***


```{r Synthetic control construct, echo=TRUE, include=TRUE}
dataprep_out <- dataprep(
  foo = prep_data_in,
  predictors = c("poverty", "income", "bmprison", "alcohol", "aidscapita", "black", "perc1519"),
  predictors.op = "mean",
  time.predictors.prior = 1990:1993,
  dependent = "State.Rate",
  unit.variable = "statefip",
  unit.names.variable = "state",
  time.variable = "year",
  treatment.identifier = 48,
  controls.identifier = c(1,2,4:6,8:13,15:42,44:47,49:51,53:56),  
  time.optimize.ssr = 1990:1993,
  time.plot = 1990:2000
)


```


***



**dataprep_out**  obtains these 4 matrices for us and we are now able to perform synthetic control analysis. Graphs below show that for the pre-treatment period 1990-1993, synthetic and treated group fit really well. Corresponding table further demonstrates that all the covariates between the treated and the synthetic cases are very similar  The shortcoming of our analysis is that we do not have a better data that goes back to at least 1985 to demonstrate the fit better. 




```{r, include=FALSE}
print(dataprep_out$X0) # check 
print(dataprep_out$Z1) # teen pregnancy rate in the pre-intervention period
synth_out <- synth(data.prep.obj = dataprep_out)
```


```{r, include=TRUE, fig.cap="Treated (Texas) Vs Synthetic Groups", fig.align="center"}
path.plot(synth_out, dataprep_out)
```



```{r, fig.cap="Table for Treated Vs Synthetic Groups"}
synth.tables = synth.tab(dataprep.res = dataprep_out,
                         synth.res = synth_out)
#names(synth.tables)

synth.tables$tab.pred[1:7,]%>%round(2)%>%kable()


```




The SCM allows us to check the relative importance of each unit and the table shows that from the first 10 units, California which is unit number 6 contributes 62.3% to the case. (P.S. analyzing all the US states contributions to the case, the rest of 37.7% is from unit number 28 which is Mississippi).




```{r, fig.cap="The Relative Importance of Each Unit", include=TRUE, fig.align="center"}
synth.tables$tab.w[1:10, ]%>%kable()
```



Gaps plot lets us visualize actual difference between the observed outcome (teen pregnancy rate in Texas with prison expansion ) and synthetic control we created for it (teen pregnancy rate in Texas without expansion). 
Thus our goal is really to find &alpha;=Y_{with_expansion_TX}+Y_{without_expansion_TX}. 


To reiterate **Y_{without_expansion_TX}** which refers to pregnancy rate before Texas prison expansion program is a synthetic control while **Y_{with_expansion_TX}** is pregnancy rate after the prison expansion which is the observed outcome. The difference is the true causal estimate of the prison expansion program of Texas on teen pregnancy rate. The graph shows that difference. It is interesting that our finding is somewhat counterintuitive. We see that the real difference between teen pregnancy rate after prison expansion and prior to it is actually positive. We hypothesized that it would be negative since there would be fewer sexual encounters between teens and young adults because of higher incarceration rate (especially higher juvenile incarceration rate). Explaining reasons behind our positive difference is beyond the scope of this research paper. 




```{r, include=TRUE, cache=TRUE, fig.cap="Actual Difference Between Teen Pregnancy rate After and Before Texas Prison Expansion", fig.align="center"}

gaps.plot(synth_out, dataprep_out, Ylab = 'Gap In Real Teen Pregnancy Birth Rate', 
         
          Xlab= 'year',
        Main = NA)
```



```{r, include=FALSE}
placebos <- generate.placebos(dataprep_out, synth_out, Sigf.ipop = 3)
```


```{r, include=TRUE, fig.align="center"}
mspe.plot(placebos, discard.extreme = TRUE, mspe.limit = 1, plot.hist = TRUE)

```


It is important to note that there are some shortcomings to our analysis. For future research, it would be beneficial to have a better data that has more covariates to predict the teen pregnancy rate. Furthermore, averaging predictors over the three-year window may not be sufficient and that extending the period will
yield a better-fitting synthetic. Unfortunately, it was difficult to find proper data to conduct better analysis and usually finding right data to conduct the SCM is one of the biggest challenge to analysts and researchers. 







\newpage



# **Conclusion**



The SCM is a quantitative method for assessing regional development policies that can complement
more qualitative methods, such as the case study approach. The method has several advantages. It is
ideal for examining a policy unique to a particular region in our case Texas prison expansion program. It makes less restrictive assumptions than
some other quantitative methods, such as difference-in-differences. The code is freely distributed and
straightforward to use. Standard output includes a list of regions and their contribution to the synthetic
control region, allowing analysts to make informed judgments about donor regions’ comparability with
the treated region. 



However, there are some limitations. Only the region
being studied should receive the treatment. There cannot be spillovers, and regions in the donor pool
cannot have engaged in similar treatments. The policy cannot have affected the region before going into
effect.  In the pre-treatment period, regions in the
donor pool must have comparable predictor variable characteristics with those of the policy region, and
those variables must have an approximately linear effect on the outcome. With this limitations in mind, conducting the SCM could be a useful tool 
in economics and public policy research. 



Our analysis of Texas prison expansion on teen pregnancy rate was a good example of the synthetic control. We found that there was actually a positive difference between teen pregnancy rate after and before Texas prison expansion. The findings were interesting and somewhat counterintuitive but it is a good starting point for further research and analysis with better data that includes more covariates and longer pre-treatment period. 





\newpage






# **Citation**





Abadie, Alberto, et al. “Synthetic Control Methods for Comparative Case Studies: Estimating the Effect of California's Tobacco Control Program.” 2007, doi:10.3386/w12831. 

<bk> <bk> <bk>

Cunningham, S. (n.d.). Causal Inference: The Mixtape. Welcome | Causal Inference. https://mixtape.scunning.com/. 

<bk> <bk> <bk>

Ye, L. P. R. (2020, January 5). Causal Inference Using Synthetic Control: The Ultimate Guide. Medium. https://towardsdatascience.com/causal-inference-using-synthetic-control-the-ultimate-guide-a622ad5cf827


